---
title: "R Data Manipulation with data.table"
author: "Wu Gong"
date: "`r Sys.Date()`"
output:
  rmdformats::readthedown:
    self_contained: true
    thumbnails: true
    code_folding: hide
    toc_depth: 6
    lightbox: true
    gallery: false
    highlight: tango
---


```{r setup, echo=FALSE, cache=FALSE,warning=FALSE,message=FALSE}
library(knitr)
library(rmdformats)
library(Wu)
## Global options
options(max.print="2000")
opts_chunk$set(echo=TRUE,
               eval=TRUE,
               cache=FALSE,
               results="asis",
               prompt=FALSE,
               tidy=FALSE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
eval_fast <- TRUE
eval_slow <- FALSE

```

# Links

[Index](index.html)



# Fill Missing Values

## Replace elements in a vector by a specific value


```{r}
x <- 1:10
list <- c(2, 4, 9)
value <- 0

replace(x, list, value)

```
## Index of observations

```{r}
dt <- data.table(
    group=c("A", "A", "B", "B", "B", "C", "C", "C", "C")
  , var1=1:9
)

dt <- dt[, index_i := .I
         ][, index_n := 1:.N
           ][, index_i_group := .I, by = .(group)
             ][, index_n_group := 1:.N, by = .(group)]

dt %>% prt()

```


## Last Observation Carry Forward (LOCF)

* data.table::nafill funtion only works for numeric variables
* Workaround for character variables by filling indexes of character variables instead <https://github.com/Rdatatable/data.table/issues/3992#issuecomment-546295277>
* type="nocb": Next Observation Carry Backward
* type="const": fill with constant

```{r}
dt <- data.table(
    n1=c(1, NA, 2, NA, NA, 3, 3, NA)
  , c1=c("A", "A", "B", NA, "C", NA, NA, NA)
)


dt <- dt[
  , n1_locf := nafill(n1, type="locf")
][, c1_locf := c1[nafill(replace(.I, is.na(c1), NA), type="locf")]
  ][, index := .I]

dt[, .(index, n1, n1_locf, c1, c1_locf)] %>% prt()
```

# Sum Multiple Columns


```{r,eval=FALSE}

is.1 <- function(x){x %in% c(1)}
sum.1 <- function(x){sum(as.numeric(is.1(x)))}
dm <- dm[, race_sum := sum(.SD)
       , .SDcols = paste0("race___", as.character(1:6))
       , by = 1:nrow(dm)]
dm <- dm[, conditions_num := sum(is.1(hypertension)
                               , is.1(diabetes)
                               , is.1(heart_disease)
                               , is.1(renal_disease)
                               , is.1(malignancies)
                               , is.1(respiratory_disease)
                                 )
       , by = 1:nrow(dm)
         ]

```

# Date and Time

```{r,eval=FALSE}

as.POSIXct(
    data$guidewire_datetime
  , format='%Y-%m-%d %H:%M'
  , tz='GMT')

as.Date(data$diag_ecg_date_time, format='%Y-%m-%d')

get_datetime <- function(text){
    as.POSIXct(
        trimws(text)
      , format = "%Y-%m-%d %H:%M"
      , tz = "GMT"
      , origin = '1970-01-01'
    )
}

get_ed_arrival <- function(text){
    as.POSIXct(
        trimws(text)
      , format = "%m/%d/%y %H%M"
      , tz = "GMT"
      , origin = '1970-01-01'
    )
}

tf <- tf[
  , datetime_ed_arrival := get_ed_arrival(ed_arrival_time_str)
][, year_ed_arrival := format(datetime_ed_arrival, "%Y")
  ][, month_ed_arrival := format(datetime_ed_arrival, "%m")
  ][, yearmonth_ed_arrival := format(datetime_ed_arrival, "%Y%m")
    ][, weekdays_ed_arrival := weekdays(datetime_ed_arrival)
      ][, weekdays_ed_arrival := factor(
              weekdays_ed_arrival
            , levels = c(
                  "Sunday"
                , "Monday"
                , "Tuesday"
                , "Wednesday"
                , "Thursday"
                , "Friday"
                , "Saturday"
              )
          )
        ][, weekdays_ed_arrival_n := as.numeric(weekdays_ed_arrival)
          ][, hour_ed_arrival := data.table::hour(datetime_ed_arrival)
          ][, minute_ed_arrival := data.table::minute(datetime_ed_arrival)
            ][, hour_ed_arrival_c := strftime(datetime_ed_arrival, format = "%H", tz = "GMT")
              ][, minute_ed_arrival_c := strftime(datetime_ed_arrival, format = "%M", tz = "GMT")][, hour_ed_arrival_n := hour_ed_arrival + minute_ed_arrival/60]


tf <- tf[, flag_bussiness_hours := case_when(
               is.na(hour_ed_arrival) ~ as.character(NA)
             , hour_ed_arrival < 8 ~ "Non-Business Hours"
             , hour_ed_arrival == 17 & minute_ed_arrival >= 1 ~ "Non-Business Hours"
             , hour_ed_arrival > 17 ~ "Non-Business Hours"
             , TRUE ~ "Business Hours"
           )
         ][, flag_bussiness_hours := factor(
                 flag_bussiness_hours
               , levels = c("Business Hours", "Non-Business Hours")
             )]

tf <- tf[
  , flag_weekday := case_when(
        weekdays_ed_arrival %in% c("Tuesday", "Wednesday", "Thursday") ~ "Weekday"
      , weekdays_ed_arrival %in% c("Saturday", "Sunday") ~ "Weekend"
      , weekdays_ed_arrival %in% c("Friday") & hour_ed_arrival == 17 & minute_ed_arrival >= 1  ~ "Weekend"
      , weekdays_ed_arrival %in% c("Friday") & hour_ed_arrival > 17  ~ "Weekend"
      , weekdays_ed_arrival %in% c("Friday") ~ "Weekday"
      , weekdays_ed_arrival %in% c("Monday") & hour_ed_arrival == 7 & minute_ed_arrival <= 59 ~ "Weekend"
      , weekdays_ed_arrival %in% c("Monday") & hour_ed_arrival < 7 ~ "Weekend"
      , weekdays_ed_arrival %in% c("Monday") ~ "Weekday"
      , TRUE ~ as.character(NA)
    )
][, flag_weekday := factor(flag_weekday, levels = c("Weekday", "Weekend"))
  ][]


fp <- fp[
    , datetime_proc_start := get_datetime(proc_start_time_str)
][, year_proc_start := format(datetime_proc_start, "%Y")
  ][, yearmonth_proc_start := format(datetime_proc_start, "%Y%m")]


vs4 <- vs4[
  , ed_arrival_to_pas_discharge := as.numeric(difftime(
        datetime_pas_discharge
      , datetime_ed_arrival
      , units = "hours"
    ))
]

vs4 <- vs4[
    , date_ed_arrival := date(datetime_ed_arrival)
][, age_in_years := as.numeric(date_ed_arrival - dob)/365.25
  ][, age_group := case_when(
          age_in_years < 18 ~ "< 18"
        , age_in_years < 65 ~ "18-64"
        , age_in_years >= 65 ~ "65 +"
        , TRUE ~ as.character(NA)
      )
    ]


```

# Computing Environment


```{r}
sessionInfo()
```
