---
title: "Heatmaps for Correlation"
author: "Wu Gong"
date: "`r Sys.Date()`"
output:
  rmdformats::readthedown:
    self_contained: true
    thumbnails: false
    code_folding: hide
    toc_depth: 6
    lightbox: true
    gallery: false
    highlight: monochrome
    css: Wu.css
---


	
```{r setup, echo=FALSE, cache=FALSE,warning=FALSE,message=FALSE}
library(knitr)
library(rmdformats)
library(Wu)
opts_chunk$set(echo=TRUE
             , cache=FALSE
             , eval=TRUE
             , prompt=FALSE
             , results="asis"
             , tidy=FALSE
             , comment=NA
             , message=FALSE
             , warning=FALSE
             , out.width = '80%'
             , class.source='klippy'
               )
eval_fast <- TRUE
eval_slow <- FALSE
klippy::klippy(position = c('top', 'left')
             , tooltip_message = 'Click to copy'
             , tooltip_success = 'Done')
```



# [Back to Index](index.html)




# Heatmap for Correlation by $R^2$

 * The correlation matrix of $R^2$ is calculated by linear regression or Cramér's V.
 * The matrix is presented using R [plotly](https://plotly.com/r/) package.

```{r, out.height='200%'}

library(data.table)
dt <- data.table(
    var1 = 1:20 + rnorm(20)
  , var2 = 1:20 + rnorm(20) * 2
  , var3 = 1:20 + rnorm(20) * 4
  , var4 = 1:20 + rnorm(20) * 8
  , var5 = 1:20 + rnorm(20) * 16
  , var6 = sample(c("A", "B"), size = 20, replace = TRUE)
  , var7 = sample(c("I", "II", "III"), size = 20, replace = TRUE)
)

Wu::heatmap(dt)


```


# Heatmap for Spearman's Correlation Coefficient

 * Use [ggcorrplot](http://www.sthda.com/english/wiki/ggcorrplot) package to present the heatmap of a correlation matrix of Spearman's $\rho$

```{r}
library(ggcorrplot)
library(data.table)

dt <- data.table(
    var1 = 1:20 + rnorm(20)
  , var2 = 1:20 + rnorm(20) * 2
  , var3 = 1:20 + rnorm(20) * 4
  , var4 = 1:20 + rnorm(20) * 8
  , var5 = 1:20 + rnorm(20) * 16
)


NVars <- c("var1"
         , "var2"
         , "var3"
         , "var4"
         , "var5"
           )

mtx <- cor(dt[, ..NVars]
         ## , na.rm=FALSE
         , use="na.or.complete"
         , method="spearman")


ggcorrplot(mtx, 
           hc.order = FALSE,
           outline.color = "white", 
           type = "lower",
           ## colors = c("red", "white", "blue"), 
           show.diag = TRUE, 
           lab = TRUE
         , tl.srt=30
         , lab_size=10
         , tl.cex=10
           , show.legend = FALSE
         ## , ggtheme = ggplot2::theme_gray
         , colors = c("#6D9EC1", "white", "#E46726")
         ## , method="circle"
           )

library(corrplot)
library(corrgram)


corrplot::corrplot(mtx)

corrgram::corrgram(mtx)

```






# treemap


```{r}


library(treemap)
data(GNI2014)
treemap(GNI2014,
       index=c("continent", "iso3"),
       vSize="population",
       vColor="GNI",
       type="value")

```


# Variable Clustering

```{r}
library(data.table)

dt <- data.table(
    var1 = 1:20 + rnorm(20)
  , var2 = 1:20 + rnorm(20) * 2
  , var3 = 1:20 + rnorm(20) * 4
  , var4 = 1:20 + rnorm(20) * 8
  , var5 = 1:20 + rnorm(20) * 16
)

library(ClustOfVar)

tree <- hclustvar(X.quanti = as.data.frame(dt))
plot(tree)

```


# Contour

```{r}
n <- 200
x <- runif(n)
x <- unique(x)
x <- sort(x)
y <- runif(n)
y <- unique(y)
y <- sort(y)

## dt <- dt[, trt := factor(trt)]


z1 <- x ^ 2 %*% t(rep(1, times = length(y)))
z2 <- rep(1, times = length(x)) %*% t(y ^ 2)
z <- z1 + z2

filled.contour(x, y, z)


```

# R sessionInfo

```{r}
sessionInfo()

```
