---
title: "R Multi-State Models"
author: "Wu Gong"
date: "`r Sys.Date()`"
output:
  rmdformats::readthedown:
    self_contained: true
    thumbnails: true
    code_folding: hide
    toc_depth: 6
    lightbox: true
    gallery: false
    highlight: tango
---


```{r setup, echo=FALSE, cache=FALSE,warning=FALSE,message=FALSE}
library(knitr)
library(rmdformats)
library(Wu)
## Global options
options(max.print="2000")
opts_chunk$set(echo=TRUE,
               eval=TRUE,
               cache=FALSE,
               results="asis",
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
eval_fast <- TRUE
eval_slow <- FALSE

```

# Multi-state model with msm package

[Multi-state modelling with R: the msm package](1 "https://cran.r-project.org/web/packages/msm/vignettes/msm-manual.pdf")

## CAV Data

Sharples et al. studied the progression of coronary allograft vasculopathy (CAV), a post-transplant deterioration of the arterial walls, using these data. Risk factors and the accuracy of the screening test were investigated using multi-state Markov and hidden Markov models.

 * 622 patients
 * Approximately each year after transplant, every patient has an angiograme when CAV could be diagnosed.
 * age: age at screen
 * dage: donor's age
 * sex: 0=male, 1=female
 * pdiag: primary diagnosis
 * cumrej: cumulative number of rejection episodes
 * firstobs: 1 = the first observation (transplant), 0 = later angiogram

L.D. Sharples, C.H. Jackson, J. Parameshwar, J. Wallwork, and S.R. Large. Diagnostic accuracy of coronary angiography and risk factors for post-heart-transplant cardiac allograft vasculopathy. Transplantation, 76(4):679682, 2003.

```{r}
library(msm)
cav %>% DT()

```

## Data Summary


```{r}
colnames(cav)

Vars <- c(
    "age"
  , "years"
  , "dage"
  , "sex"
  , "pdiag"
  , "cumrej"
  , "state"
)

FactorVars <- c(
    "sex"
  , "pdiag"
  , "cumrej"
  , "state"
)

t <- Table1n(obj = cav, Vars = Vars, FactorVars = FactorVars)

t %>% prt()

```

## Simple bidirectional model

 * The day of death is assumed to be recorded exactly. The state 4 is death, deathexact=4
 * If the data has two death states 4 and 5 with two competing risks, deathexact=c(4,5)
 * State transition table: from state (in row) to state (in column)
 * Transition intensity matrix Q: zeros in the matrix
 * obstype: a vector specifying the observation scheme for each row of the data
     + =1: a snapthot of the process, the states are unkown between observation times
	 + =2: an exact transition time, the state at the previous observation retained untile the current observation
	 + =3: an exact transition time, but the state at the instant before entering this state is unknown.
	 + exacttime=TRUE: all observations are of obstype 2.
	 + deathexact=death.states specifies that all observations of death.states are of type 3.
	 + deathexact=TRUE specifies all observations in the final absorbing state are of type 3.

```{r}
library(msm)


t <- statetable.msm(state, PTNUM, data=cav)

t %>% as.data.frame.matrix() %>% prt(caption="Transition Frequencies")

Q <- rbind ( c(0, 0.25, 0, 0.25),
            c(0.166, 0, 0.166, 0.166),
            c(0, 0.25, 0, 0.25),
            c(0, 0, 0, 0) )


Q %>% prt(caption="Initial Transition Intensity")

cav.msm <- msm(state ~ years
             , subject = PTNUM
             , data = cav
             , qmatrix = Q
             , deathexact = 4)

pv <- summary(cav.msm)$prevalences

pv$Observed %>% prt(caption="Observed Frequencies")
pv$`Observed percentages` %>% prt(caption="Observed percentages")

pv$Expected %>% prt(caption="Expected Frequencies")
pv$`Expected percentages` %>% prt(caption="Expected percentages")



P <- pmatrix.msm(cav.msm, t=1, t1=0, ci="normal")

as.data.frame.matrix(P$estimates) %>% prt(caption="Transition Probabilities from time 0 to time 1")


prv <- prevalence.msm(
    cav.msm
  , times=5
  , timezero=NULL
  , initstates=NULL
  , covariates="population"
)


prv[["Expected percentages"]] %>% prt(caption="predicted prevalence percentage at time 5")

```

## Model with Covariates

```{r}
cav.msm.1 <- msm(state ~ years
             , subject = PTNUM
             , data = cav
             , qmatrix = Q
             , deathexact = 4
             , covariates = ~ sex
               )

h1 <- hazard.msm(cav.msm.1)

h1$sex %>% prt(caption="Hazard Ratio of sex 1 against 0")

sj1 <- sojourn.msm(cav.msm.1, covariate=list(sex=1), ci="norm")
sj0 <- sojourn.msm(cav.msm.1, covariate=list(sex=0), ci="norm")

sj1 %>% prt(caption="Estimated Time for Each State sex=1")
sj0 %>% prt(caption="Estimated Time for Each State sex=0")

plot(cav.msm)
plot(cav.msm.1)
plotprog.msm(state ~ years, subject = PTNUM, data=cav)

```


## Model Comparison


```{r}
cav.msm <- msm(state ~ years
             , subject = PTNUM
             , data = cav
             , qmatrix = Q
             , deathexact = 4)

cav.msm.1 <- msm(state ~ years
             , subject = PTNUM
             , data = cav
             , qmatrix = Q
             , deathexact = 4
             , covariates = ~ sex
               )

t <- lrtest.msm(cav.msm, cav.msm.1)

t %>% prt(caption="Likelihood Ratio Test: Null model vs sex model"
          , col.names=attr(t, "dimnames")[[2]]
          )

```

# Computing Environment


```{r}
sessionInfo()
```
