---
title: "Power Analysis"
author: "Wu Gong"
date: "`r Sys.Date()`"
output:
  rmdformats::readthedown:
    self_contained: true
    thumbnails: true
    code_folding: hide
    toc_depth: 6
    lightbox: true
    gallery: false
    highlight: tango
---


```{r setup, echo=FALSE, cache=FALSE,warning=FALSE,message=FALSE}
library(knitr)
library(rmdformats)
library(Wu)
## Global options
options(max.print="2000")
opts_chunk$set(echo=TRUE,
               eval=TRUE,
               cache=FALSE,
               results="asis",
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
eval_fast <- TRUE
eval_slow <- FALSE

```



# Non-inferiority Cluster Randomised Trial

Richard J. Hayes & Lawrence H. Moulton. Cluster Randomised Trials, Second Edition, 2017, P. 156

* This formula uses alpha value of 0.05 not 0.025, to ensure a standard 95% confidence interval on delta would excluded the -0.05 margin with 80% probability.
* C: number of clusters
* m: number of subjects per cluster
* delta = pi_1 - pi_0: true difference between two arms
* pi_1: proportion on the treatment arm
* pi_0: proportion on the control arm
* beta:
* alpha:
* k: coefficient of variance

```{r}

C <- 104 / 3
m <- 30
delta <- -0.05
pi_0 <- 0.15
k <- 0.25
alpha <- 0.025
z_alpha <- qnorm(1 - alpha)

c <- 1 + (z_alpha + z_beta) ^ 2 * (2 * pi_0 * (1 - pi_0) / m + 2 * k ^ 2 * pi_0 ^ 2) / (delta ^ 2)

xx <- (2 * pi_0 * (1 - pi_0) / m + 2 * (k ^ 2) * (pi_0 ^ 2)) / (delta ^ 2)
z_beta <- sqrt((C - 1) / xx) - z_alpha

pnorm(z_beta)

```



# Computing Environment


```{r}
sessionInfo()
```
