---
title: "R Markdown with Tabs"
author: "Wu Gong"
date: " `r Sys.Date() ` "
output:
  html_document:
    toc: no
    toc_depth: 4
    toc_float:
      collapsed: true
    code_folding: hide
    theme: readable
    number_sections: false
    fig_caption: true
---

```{r setup, include = FALSE}
rm(list = ls())
library(Wu)
knitr::opts_chunk$set(
  echo = FALSE
  , message = FALSE
  , tidy = TRUE
  , results = "asis"
  , warning = FALSE
  , fig.width = 12
  , fig.height = 8
)

options(knitr.kable.NA = "")

theme_update(
    axis.text = element_text(size = 10)
  , axis.title = element_text(size = 10)
)
```
<style>
.author{
    display: none;
}
.title{
    display: none;
}
.date{
    display: none;
}
.main-container {
  max-width: 2400px;
  margin-left: auto;
  margin-right: auto;
}
@media print {
    .tab-pane {
        display: block !important;
        opacity: 1;
    }
}
</style>

# {.tabset .tabset-fade .tabset-pills}

## Level I Tab A

* A R markdown template that outputs a structure of hiercharical tabs


## Tab B

```{r}

plot(1:10)

```


## Tab C {.tabset .tabset-fade}

```{r}
library(Wu)
dt <- data.table(var1 = 1:10, var2 = 11:20)
dt %>% prt()



```


## Tab D {.tabset .tabset-fade}

### Tab I

```{r}
plot(1:10, (1:10) ^ 2)
```

### Tab II

```{r}
plot(1:10, (1:10) ^ 0.5)
```

### Tab III

```{r}
plot(1:10, (1:10) * 2)
```



## About

[Wu Gong](https://ghowoo.github.io/)

```{r}
print(Sys.Date())
```

```{r}
sessionInfo()
```
