---
title: "Time-To-Event Analysis"
author: "Wu Gong"
date: "`r Sys.Date()`"
output:
  rmdformats::readthedown:
    self_contained: true
    thumbnails: false
    code_folding: hide
    toc_depth: 6
    lightbox: true
    gallery: false
    highlight: monochrome
    css: Wu.css
---


```{r setup, echo=FALSE, cache=FALSE,warning=FALSE,message=FALSE}
library(knitr)
library(rmdformats)
library(Wu)
opts_chunk$set(echo=TRUE,
               cache=FALSE,
               eval=TRUE,
               prompt=FALSE,
               results="asis",
               tidy=FALSE,
               comment=NA,
               message=FALSE,
               warning=FALSE,
               out.width = '80%')
eval_fast <- TRUE
eval_slow <- FALSE

```

# Weibull AFT Regression in R

[Sarah R Haile: Weibull AFT Regression Functions in R](https://cran.r-project.org/web/packages/SurvRegCensCov/vignettes/weibull.pdf "Sarah Haile: Weibull AFT Regression in R")


Accelerated Failure Time model:

$$ logT = Y = \mu + \alpha^{T}z + \sigma W $$

W has the extreme value distribution.

[Extreme Value Distribution](https://www.itl.nist.gov/div898/handbook/apr/section1/apr163.htm)

* Extreme value distributions arise as limiting distributions for maximums or minimums (extreme values) of a sample of independent, identically distributed random variables, as the sample size increases.
* In the context of reliability modeling, extreme value distributions for the minimum are frequently encountered. For example, if a system consists of n identical components in series, and the system fails when the first of these components fails, then system failure times are the minimum of n random component failure times.
* Extreme value theory says that, independent of the choice of component model, the system model will approach a Weibull as n becomes large.
* The same reasoning can also be applied at a component level, if the component failure occurs when the first of many similar competing failure processes reaches a critical level.



$$ \gamma = 1 / \sigma $$

$$ \lambda =  exp(-\mu/\sigma)$$
$$ \beta = -\alpha/\sigma $$

The Hazard function of the Weibull model, the first part is the baseline hazard.

$$ h(x|z) = (\gamma \lambda t^{\gamma - 1} ) exp(\beta^{T} z) $$


[Weibull](https://stat.ethz.ch/education/semesters/ss2011/seminar/contents/handout_9.pdf)

* $ \gamma $ is a shape parameter
* $ \gamma > 1 $ the hazard increases
* $ \gamma = 1 $ the hazard is constant, Weibull reduces to exponential distribution
* $ \gamma < 1 $ the hazard decreases




Hazard Ratio:
$$ exp(\beta_i) $$

Event Time Ratio (ETR): the ratio quantifies the relative difference in time it takes to achieve the pth percentile between two lelvels of a covariate
$$ exp(\alpha_i) = exp(-\beta_i/\gamma) $$

```{r}
library(survival)
library(KMsurv)
library(SurvRegCensCov)
data(larynx)

m <- WeibullReg(Surv(time, death) ~ factor(stage) + age, data=larynx)

formula(m)
m$coef


```

# Weibull Distribution


```{r}
scale <- 1
shape <- 1
x <- seq(0, 5, by=0.1)
yd <- dweibull(x, shape=shape, scale=scale)

plot(x, yd)


```



# Simulating Survival Data

<https://stats.stackexchange.com/questions/135124/how-to-create-a-toy-survival-time-to-event-data-with-right-censoring>


```{r}
# baseline hazard: Weibull

# N = sample size
# lambda = scale parameter in h0()
# rho = shape parameter in h0()
# beta = fixed effect parameter
# rateC = rate parameter of the exponential distribution of C


simulWeib <- function(N, lambda, rho, beta, rateC)
{
  # covariate --> N Bernoulli trials
  x <- sample(x=c(0, 1), size=N, replace=TRUE, prob=c(0.5, 0.5))

  # Weibull latent event times
  v <- runif(n=N)
  Tlat <- (- log(v) / (lambda * exp(x * beta)))^(1 / rho)

  # censoring times
  C <- rexp(n=N, rate=rateC)

  # follow-up times and event indicators
  time <- pmin(Tlat, C)
  status <- as.numeric(Tlat <= C)

  # data set
  data.frame(id=1:N,
             time=time,
             status=status,
             x=x)
}

```

# Weibull with R survreg Package

* Set scale=0, the scale is estimated


```{r}
survreg(Surv(futime, fustat) ~ ecog.ps + rx
      , ovarian
      , dist='weibull'
      , scale=1
        )

survreg(Surv(futime, fustat) ~ ecog.ps + rx
      , ovarian
      , dist='weibull'
      , scale= 0
        )

```

# AFT Model

[AFT Model](https://stat.ethz.ch/education/semesters/ss2011/seminar/contents/handout_9.pdf)

$$ log(T) = \alpha_0 + \alpha_1 X + \epsilon$$

* log(T) = extreme value distribution: Exponential
* log(T) = extreme value distribution: Weibull, which has an additional parameter that scales $\epsilon$
* log(T) = logistic: log-logistic
* long(T) = normal: log-normal

* Accelerated time failure assumption: The probability of a dog surviving past t years is equal to a human surviving past 7*t years.
* AFT Models describe stretching out or contraction of survival time as a function of predictor variables.
* AFT assumption: S is survival, $\gamma$ is acceleration factor
$$ S_{non-treatment} = S_{treatment}(\gamma t)$$
$$ \gamma T_{non-treatment} = T_{treatment}$$
* $\gamma > 1$ exposure benefits survival




# Discrete Time Proportional Odds

## Data Simulation

* Assume average daily death rate of 5%
* Individual baseline death rate has a random effect has standard deviation of the half of the log odds
* The treatment effect of odds ratio 0.5
* Subjects are observed for 28 days


```{r}
library(Wu)

set.seed(123456)
n <- 1000
n0 <- 500
n1 <- n - n0

r_0 <- 0.05
odds_0 <- r_0 / (1 - r_0)
sd <- abs(log(odds_0) / 2)
odds_0s <- exp(rnorm(n0, log(odds_0), sd))
r_0s <- odds_0s / (1 + odds_0s)

or <- 0.5
odds_1 <- odds_0 * or
r_1 <- odds_1 / (1 + odds_1)
odds_1s <- exp(rnorm(n1, log(odds_1), sd))
r_1s <- odds_1s / (1 + odds_1s)

days <- 28
for(i in 1:days){
    print(i)
    event_i <- c(rbinom(n0, 1, r_0s), rbinom(n1, 1, r_1s))
    if(i == 1){
        t <- data.table(
            day=i
          , id=1:n
          , trt=rep(c(0, 1), each=c(n0, n1))
          , event=event_i
        )
        rtn <- copy(t)
    }else{
        event_i[t$event %in% c(1, NA)] <- NA
        t <- data.table(
            day=i
          , id=1:n
          , trt=rep(c(0, 1), each=c(n0, n1))
          , event=event_i
        )
        rtn <- rbind(rtn, t)
    }
}

dim(rtn) %>% prt()
rtn[, .N, by = .(day)][order(day)] %>% prt()
rtn[, .N, by = .(day, trt)][order(day)] %>% prt()
rtn[, .N, by = .(day, event)][order(day, event)] %>% prt()

dt <- copy(rtn)[!is.na(event)]
dim(dt)

dt <- dt[order(id, -day)
         ][, rn := 1:.N, by = .(id)
           ][rn == 1
             ][, rn := NULL]
dt[, .N, by = .(day, event)][order(day, event)]
dt <- dt[, censor := case_when(event == 0 ~ 1, TRUE ~ 0)]

```

## Kaplan Meier Plot

```{r}
library(survival)
library(survminer)

f <- survfit(Surv(day, event) ~ trt, data = dt)
plot(f)

ggsurvplot(f)

```

## Longitudinal Mixed-Effect Model R::lme4

```{r}
library(lme4)
dt <- copy(rtn)
dt <- dt[!is.na(event)]

t <- get_ors(outcome="event", predictor="trt", data=dt, digits = 3)
t

m <- glmer(event ~ trt + (1 | id)
         , data=dt
         , family = binomial(link=logit)
           )


summary(m)

exp(fixef(m))



```

## Bayesian Longitudinal Mixed-Effect Model R::brms

```{r,eval=FALSE}
library(brms)
dt <- copy(rtn)
dt <- dt[!is.na(event)]

m <- brm(event ~ trt + (1 | id)
       , data=dt
       , family = binomial(link=logit)
       , chain=5
       , cores=5
       , warmup=200
       , iter = 400
         )

summary(m)
exp(fixef(m))
ranef(m)

```

```{r,eval=FALSE}

library(rmsb)

n_iter <- 8000
n_warmup <- n_iter / 2
t1 <- Sys.time()
library(rmsb)
dd <- datadist(dt)
options(datadist='dd')
file.remove("discrete_death_blrm.RDS")
m <- blrm(daily_death ~ age_group +
              rcs(day_assessment_icu, 5) +
              days_admit2icu3 +
              sex_2.factor +
              race_ethnicity +
              com_astha +
              com_coronary +
              com_immunosuppresion +
              com_diabetes +
              com_hypertension +
              com_stroke +
              com_renal +
              com_copd +
              com_heartfailure +
              com_obesity +
              cluster(record_id) +
              redcap_data_access_group
        , data=dt
        , iter=n_iter
        , warmup=n_warmup
        , refresh = 500
        , progress = "discrete_death_blrm_progess.log"
        , chain=4
        , cores=4
        , show_messages=TRUE
        , file="discrete_death_blrm.RDS"
          )
t2 <- Sys.time()
print(t2 - t1)
st <- blrmStats(m, ns=n_iter * 4 / 2)
## get concordance probabiilty etc.
saveRDS(st, file="discrete_death_blrmStats.RDS")




m <- readRDS(file="discrete_death_blrm.RDS")
smry <- summary(m)
rNames <- attr(smry, "dimnames")[[1]]
cNames <- attr(smry, "dimnames")[[2]]
smry <- as.data.table(smry)
colnames(smry) <- cNames
smry$coef_name <- rNames

smry[, c(9, 1, 2, 4, 6, 7)] %>%
    kable(caption="Coefficients of the Model (Odds Ratios)"
      , digits=2
      , align="rrrrrrrrrrrrrrrr"
        ) %>% styling()

```


## Bayesian Longitudinal Mixed-Effect Logistic R::rmsb

```{r, eval=FALSE}

t1 <- Sys.time()
library(rmsb)
dd <- datadist(dt)
options(datadist='dd')
file.remove("discrete_death_blrm.RDS")
m <- blrm(event_death ~ age_group +
              rcs(day_assessment_icu, 5) +
              days_admit2icu3 +
              sex_2.factor +
              race_ethnicity +
              com_astha +
              com_coronary +
              com_immunosuppresion +
              com_diabetes +
              com_hypertension +
              com_stroke +
              com_renal +
              com_copd +
              com_heartfailure +
              com_obesity +
              cluster(record_id) +
              redcap_data_access_group
        , data=dt
        , iter=2000
        , warmup=1000
        , refresh = 500
        , progress = "discrete_death_blrm_progess.log"
        , chain=4
        , cores=4
        , show_messages=TRUE
        , file="discrete_death_blrm.RDS"
          )
t2 <- Sys.time()
print(t2 - t1)
st <- blrmStats(m, ns=4000)
## get concordance probabiilty etc.
saveRDS(st, file="discrete_death_blrmStats.RDS")


```
# Multi-state Transition Model

Christopher Jackcon: Multi-state modelling with R: the msm package
https://cran.r-project.org/web/packages/msm/vignettes/msm-manual.pdf


## Basic Setting

```{r}
library(msm)

cav[1:21, ] %>% prt()

statetable.msm(state, PTNUM, data=cav) %>% prt()

Q <- rbind ( c(0, 0.25, 0, 0.25),
            c(0.166, 0, 0.166, 0.166),
            c(0, 0.25, 0, 0.25),
            c(0, 0, 0, 0) )

Q.crude <- crudeinits.msm(state ~ years, PTNUM, data=cav,
                            qmatrix=Q)
```

## Simple bidirectional model

 * The sizes of the confidence intervals for some of the hazard ratios suggest there is no information in the data about the corresponding covariate effects.


```{r}
cav.msm <- msm( state ~ years, subject=PTNUM, data = cav, qmatrix = Q, deathexact = 4)
t <- qmatrix.msm(cav.msm)

t$estimates %>% prt()

cavsex.msm <- msm( state ~ years, subject=PTNUM, data = cav, qmatrix = Q, deathexact = 4, covariates = ~ sex)

qmatrix.msm(cavsex.msm)$estimates %>% prt()

qmatrix.msm(cavsex.msm, covariates=list(sex=0))$estimates %>% prt()

```

## Transition-specific covariate

 * The transition rate from state 1 to state 2, and the rate from state 1 to state 4 are each modelled on sex as a covariate, but no other intensities have covariates on them.


```{r}

cavsex2.msm <- msm( state ~ years
                , subject=PTNUM
                , data = cav
                , qmatrix = Q
                , deathexact = 4
                , covariates = list("1-2" = ~ sex, "1-4" = ~sex)
                  )
qmatrix.msm(cavsex2.msm)$estimates %>% prt()

```

## Constrained covariate effects

 * This contrains the effect of sex to be equal for the progression rates q12, q23, and recovery rates q21, q32.
 * The index on the transition intensities, ordered by rows, could be used as constraint indicators. There are totally 7 intensities and 7 hazard ratios for covariate sex. list(sex=c(1,2,3,4,5,6,7)) means all hazard ratios are distinct, and list(sex=c(1,2,3,4,5,6,1) means the first and 7th hazard ratios are same.


```{r}
cav3.msm <- msm( state ~ years, subject=PTNUM
              , data = cav, qmatrix = Q, deathexact = 4
              , covariates = ~ sex
              , constraint = list(sex=c(1,2,3,4, 5, 6, 1)) )
qmatrix.msm(cav3.msm)$estimates %>% prt()
```

# Computing Environment


```{r}
sessionInfo()

```
